---
title: "admr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{admr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Example of using `admr` with `rxode2`
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Importing the data
Import the `admr` and other packages:
```{r setup}
library(admr)
library(rxode2)
library(nlmixr2)
library(dplyr)
library(tidyr)
library(mnorm)
```


```{r}
head(examplomycin)
```

```{r}
examplomycin_wide <- examplomycin %>%
  filter(EVID != 101) %>%
  dplyr::select(ID, TIME, DV) %>%  # Select only the relevant columns
  pivot_wider(names_from = TIME, values_from = DV) %>%
  dplyr::select(-c(1))

examplomycin_aggregated <- examplomycin_wide %>%
  admr::meancov()

head(examplomycin_aggregated)
```


## Defining the RxODE model
We will define a two-compartment model for the examplomycin dataset. The model has the following differential equations:
```{r}
rxModel <- RxODE({
  # Parameters
  ke = cl / v1             # Elimination rate constant
  k12 = q / v1             # Rate constant for central to peripheral transfer
  k21 = q / v2             # Rate constant for peripheral to central transfer
  
  # Differential equations for drug amount in compartments
  d/dt(depot)    = -ka * depot
  d/dt(central)  = ka * depot - ke * central - k12 * central + k21 * peripheral
  d/dt(peripheral) = k12 * central - k21 * peripheral
  
  # Concentration in the central compartment
  cp = central / v1
})
```

A solved model is available for the two-compartment model:
```{r}
rxModel <- RxODE({
  cp = linCmt(    # Solved one- or two-compartment model
    cl,           # Clearance
    v1,           # Volume of the central compartment
    v2,           # Volume of the peripheral compartment
    q,            # Inter-compartmental clearance
    ka            # Absorption rate constant
  )
})
```

## Defining the prediction function
```{r}
predder <- function(xt, theta_i, dose = 100) {
  if (!is.matrix(theta_i) || nrow(theta_i) == 1) {
    # Create the event table for dosing and sampling
    ev <- rxode2::eventTable()
    ev$add.dosing(dose = dose, nbr.doses = 1, start.time = 0)  # Single dose
    ev$add.sampling(xt)  # Time points to simulate
    
    # Solve the RxODE model
    out <- rxSolve(rxModel, theta_i, events = ev, cores = 10)
    
    # Extract the predicted concentrations (cp)
    return(out$cp)
    
  } else {
    # Multiple individuals case
    n_individuals <- nrow(theta_i)

    # Create the event table for dosing and sampling
    ev <- eventTable(amount.units="mg", time.units="hours")
    ev$add.dosing(dose = dose, nbr.doses = 1, start.time = 0)  # Assume same dose
    ev$add.sampling(xt)  # Time points to simulate
    
    # Solve the RxODE model for all individuals
    out <- rxSolve(rxModel, params = theta_i, events = ev, cores = 10)
    
    # Extract the predicted concentrations (cp) and format them as a matrix
    cp_matrix <- matrix(out$cp, nrow = n_individuals, ncol = length(xt), byrow = TRUE)
    
    # Return matrix of predicted concentrations
    return(cp_matrix)
  }
}

twocomp_g <- function(beta,bi=rep(0,length(beta)),ai) {
  beta*exp(bi)
}
```

Specify the options for the `admr` function:
```{r}
opts <- genopts(time=c(.1,.25,.5,1,2,3,5,8,12),
                init=list(beta=c(cl = 5, v1 = 10, v2 = 30,q = 10, ka = 1),Omega=omegas(.09,0,5),Sigma_prop=0.04),
                nsim=2500,n=250, g=twocomp_g,
                adist=NULL, interact=TRUE,fo_appr=FALSE,biseq=NA,omega_expansion=1.2,
                p_thetai=function(p,origbeta,bi) {
                  dmnorm(bi,mean=log(p$beta/origbeta),sigma=p$Omega,log=TRUE)$den
                },
                f=predder)


```


Fit the model to the examplomycin dataset:
```{r}
fit <- fitEM(opts, examplomycin_aggregated, chains = 5)
print(fit)
```

Plot the results:
```{r}
plot(fit)
```

